---
title: "Multiple Regression"

author:
  - name: António Duarte
    email: aduarte@ipb.pt
    
format:
  revealjs:
    fig-asp: 0.7
---

## Data preparation

```{r}
#| echo: true

data(mtcars)

# convert to factor
mtcars <- transform( mtcars,
  am = factor(am, labels = c("auto", "manual")),
  vs = factor(vs, labels = c("V", "S")),
  cyl = factor(cyl),
  gear = factor(gear)
)
```

Variables `cyl` and `gear` were coded as factors to allow non linar relationships and because they more or less define a engine category.

Variable `carb` was not factorized because some categories have very low counts (`r table(mtcars$carb)`).

## Some models

```{r}
#| echo: true

# model with 2 vars
m1 <- lm(mpg ~ wt + hp, data = mtcars)

# same model without constant term (line through origin)
m2 <- lm(mpg ~ wt + hp - 1, data = mtcars)
m2 <- lm(mpg ~ 0 + wt + hp, data = mtcars) # the same

# model with all vars
m3 <- lm(mpg ~ ., data = mtcars)

# model with all vars except `gear`
m4 <- lm(mpg ~ . - gear, data = mtcars)
```


## Predictions

```{r}
#| echo: true

nd = data.frame(
  wt = c(2, 3, 4),
  hp = c(100, 120, 140)
)

predict(m1, newdata = nd, interval = "prediction")
```

`newdata` must point to a data frame containing variables with the same names as specified in the model formula

## Model with factors

```{r}
#| echo: true

(m5 <- lm(mpg ~ wt + vs, data = mtcars))

nd = data.frame(
  wt = c(1, 2, 1, 2),
  vs = c("V", "V", "S", "S")
)

predict(m5, newdata = nd)
```



## Model plot (`m5`)

```{r}
#| fig-align: center

palette("Okabe-Ito")

plot(mpg ~ wt, data = mtcars, col = vs)

# auto cars
abline(
  a = m5$coefficients["(Intercept)"],
  b = m5$coefficients["wt"],
  col = 1
)

# manual cars
abline(
  a = m5$coefficients["(Intercept)"] + m5$coefficients["vsS"],
  b = m5$coefficients["wt"],
  col = 2
)

#legend
legend(
  "topright",
  legend = c("vs = V", "vs = S"),
  col = 1:2,
  lty = "solid",
  bty = "n")
```

Slope is the same for V and S cars, fit can be improved


## Model with interactions

```{r}
#| echo: true

(m6 <- lm(mpg ~ wt + vs + wt:vs, data = mtcars))
(m6 <- lm(mpg ~ wt * vs, data = mtcars)) # the same

predict(m6, newdata = nd)
```

## Model plot (`m6`)

```{r}
#| fig-align: center

plot(mpg ~ wt, data = mtcars, col = vs)

# auto cars
abline(
  a = m5$coefficients["(Intercept)"],
  b = m5$coefficients["wt"],
  col = 1
)

# manual cars
abline(
  a = m6$coefficients["(Intercept)"] + m6$coefficients["vsS"],
  b = m6$coefficients["wt"] + m6$coefficients["wt:vsS"],
  col = 2
)

#legend
legend(
  "topright",
  legend = c("vs = V", "vs = S"),
  col = 1:2,
  lty = "solid",
  bty = "n")
```

With interactions, fit is improved, equivalent of fitting 2 simple models, one for V and another for S


## Variable Selection (stepwise methods)

```{r}
#| results: hide
#| echo: true

m0 <- lm(mpg ~ 1, data = mtcars) # no variables
m_all <- lm(mpg ~ ., data = mtcars) # all variables

# Forward procedure
m7 <- step(
  m0,                         # starting model
  scope = formula(m_all),     # variables that can be added
  direction = "both",         # allowing removals (default)
  test = "F"                  # show F test
)

# Backward procedure
m8 <- step(m_all, direction = "both")
```

These procedure tends to overfit the models, care is needed


## Avoiding overfitting

```{r}
#| results: hide
#| echo: true
#| 

data("penguins")

# remove rows with missing data
penguins <- na.omit(penguins)

# divide data into test and train (about 80% of data)
set.seed(2026)
train_rows <- sample(nrow(penguins), 0.8 * nrow(penguins))
train <- penguins[train_rows, ]
test <- penguins[-train_rows, ]

# estimate the model aka train the model
m0 <- lm(body_mass ~ 1, data = train)
m_all <- lm(body_mass ~ ., data = train)

mf <- step(m0, scope = formula(m_all)) # forward
mi <- update(mf, . ~ . + species:sex) # add interaction
```


## Evaluating predictions {.smaller}

```{r}
#| echo: true

# make predictions
pred_mf <- predict(mf, newdata = test)
pred_mi <- predict(mi, newdata = test)
```

```{r}
#| echo: true
#| layout-ncol: 2

# for computing metrics
library(Metrics)

# mean absolute error
mae(pred_mf, test$body_mass)
mae(pred_mi, test$body_mass)

# mean absolute percent error
mape(pred_mf, test$body_mass)
mape(pred_mi, test$body_mass)

# mean squared error
mse(pred_mf, test$body_mass)
mse(pred_mi, test$body_mass)

# root mean squared error
rmse(pred_mf, test$body_mass)
rmse(pred_mi, test$body_mass)
```

Model with interaction has slightly  better metrics (expected, because has more parameters).

## Prediction plot

```{r}
#| fig-align: center
#| layout-ncol: 2


plot(test$body_mass, pred_mf, col = test$species, pch=as.numeric(test$sex))
abline(a=0, b=1, lty = "dashed")

legend(
  "topleft",
  legend = c(levels(test$species), "", levels(test$sex)),
  col = c(1:3, NA, 1, 1),
  pch = c(1, 1, 1, NA, as.numeric(test$sex)),
  bty = "n"
)

plot(test$body_mass, pred_mi, col = test$species, pch=as.numeric(test$sex))
abline(a=0, b=1, lty = "dashed")
```

Models seem very similar
